---
title: "Basic Example"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Basic Example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Intro

A good place to start is by browsing datasets on
[opendata.scot](https://opendata.scot/). You can filter and search to
find datasets of interests. The `openscotland` R package can then
download and save the datasets you are interested in. This helps to
quickly start analysis without needing to work out how to save, organise
and import data in R.

# Search

The `opendatascotland` package has a basic search function for finding
available datasets and viewing their metadata. Use `ods_search()`
function to view metadata for all datasets or filter by matching search
terms in the dataset's title.

```{r search, warning=FALSE, echo=FALSE, message=FALSE}
library(opendatascotland)
# View all available datasets and associated metadata
all_datasets <- ods_search()

# Search dataset titles containing matching terms (case insensitive)
single_query <- ods_search("Number of bikes")

# Search multiple terms
multi_query <- ods_search(c("Bins", "Number of bikes"))
```

```r
library(opendatascotland)

# View all available datasets and associated metadata
all_datasets <- ods_search()

# Search dataset titles containing matching terms (case insensitive)
single_query <- ods_search("Number of bikes")

# Search multiple terms
multi_query <- ods_search(c("Bins", "Number of bikes"))
#> The cached list of datasets from opendata.scot was last downloaded on 2023-10-09
head(multi_query)
#> # A tibble: 6 × 11
#>   unique_id            title organization notes category url   resources licence
#>   <chr>                <chr> <chr>        <chr> <list>   <chr> <list>    <chr>  
#> 1 Litter_Bins_Aberdee… Litt… Aberdeen Ci… "<di… <chr>    /dat… <df>      UK Ope…
#> 2 Communal_Bins_City_… Comm… City of Edi… "<p>… <chr>    /dat… <df>      UK Ope…
#> 3 Grit_Bins_City_of_E… Grit… City of Edi… "<p>… <chr>    /dat… <df>      UK Ope…
#> 4 Salt_Bins_Dumfries_… Salt… Dumfries an… "<p>… <chr>    /dat… <df>      UK Ope…
#> 5 Roads_-_winter_main… Road… Stirling Co… "<p>… <chr>    /dat… <df>      UK Ope…
#> 6 Number_of_bikes_ava… Numb… Cycling Sco… "<p>… <chr>    /dat… <df>      UK Ope…
#> # ℹ 3 more variables: date_created <chr>, date_updated <chr>, org_type <chr>
```
Note, search term is case-insensitive but word order must be correct
(there is no 'fuzzy' matching).

# Download

Currently, only datasets available in `.csv`, `json` or `.geojson` can
be downloaded. These formats cover the majority of data available. You
will be warned if a dataset can't be downloaded.

To download data, you can either download the metadata using
`ods_search()`.

```{r}
query <- ods_search("bins")
```

Then pass that data frame to `ods_get()`.

```{r, echo=FALSE, message=FALSE}
data <- ods_get(query, ask = FALSE)
```

```r
data <- ods_get(query)
```

***Or*** use the `search` argument in `ods_get(search="my search term")`
to search and download any matching datasets in one step.

```{r, echo=FALSE, message=FALSE}
data <- ods_get(search = "bins", ask = FALSE)
```

```r
data <- ods_get(search = "bins")
```

By default, you will be asked if you want to save the data locally on
the first download. Optionally, you can refresh the data or avoid being
asked to save data by setting the `refresh` or `ask` arguments.

``` r
data <- ods_get(search = "Number of bikes", refresh = TRUE, ask = FALSE)
```

# Plot

Once you have downloaded the dataset(s), you may wish to plot or map the
data. Here's a short example of how plotting can be done.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
data <- ods_get(search = "Recycling Points", ask = FALSE)
```

``` r
data <- ods_get(search = "Recycling Point Locations")
```

The `ods_get()` function returned a named list of data frames - lets
select the one we want by name:

```{r}
recycling_points <- data$Recycling_Points_Aberdeen_City_Council
```

Or alternatively select the data frame using index/position of
the name in the list using `[[]]` notation.

```{r}
names(data)
recycling_points <- data[[2]]
```

Geojson datasets are automating converted to [simple
feature](https://r-spatial.github.io/sf/) 'sf' data. As we can see in
the example the data frame is classed as "sf" which means spatial /
geometry attributes are baked in.

```{r}
class(recycling_points)
```

You can see a `geometry` variable which contains the spatial
co-ordinates.

```{r}
head(recycling_points, 3)
```

This allows the `plot()` function to automatically plot
the coordinates.

```{r}
plot(recycling_points$geometry, col = as.factor(recycling_points$TEXTILE_YN))
```

## JSON

Some datasets are only available in JSON or CSV formats, unlike formats such as
GeoJSON, these do not automatically have spatial geometry. For example, the
Glasgow cycling counts data does have `latitude` and `longitude` variables but
is downloaded as JSON and therefore does not have spatial geometry added by
default. This type of data is downloaded as a data frame without a `geometry`
column.

Let's download the Glasgow cycle counts data which is only available in
JSON format.

```{r, echo=FALSE, message=FALSE}
cycle_count <- ods_get(
  search =
    "Glasgow City Council - Daily cycling counts from automatic cycling",
  ask = FALSE,
  refresh = TRUE
)
```

``` r
cycle_count <- ods_get(search = "lasgow City Council - Daily cycling counts from automatic cycling counters")
```

```{r}
cycle_count <- cycle_count[[1]]
head(cycle_count, 4)
```

We can see the cycle count JSON has been converted into a 'flat'
tabulated data frame. This data frame can now be plotted as a chart, in
the following example, let's use the `ggplot2` plotting library to
create a graph. We'll display counts over time for each location.

```{r, fig.width=5}
library(ggplot2)
library(scales)
# Convert character to date (to display date time correctly)
cycle_count$Date <- as.Date(substr(cycle_count$startTime, 1, 10))
cycle_count <- cycle_count[cycle_count$Date > as.Date("2022-12-31"), ]

# Plot
ggplot(cycle_count, aes(x = Date, y = count, colour = location)) +
  geom_line() +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_x_date(date_labels = "%b-%Y") +
  scale_y_continuous(labels = comma)
```

This is a bit noisy so lets sum the counts for all locations to see
total cycle counts.

```{r, message=FALSE, fig.width=5}
library(ggplot2)
library(dplyr)
library(magrittr)

# Group by date and sum the count
sum_cycle_count <- cycle_count %>%
  group_by(Date) %>%
  summarise(`Total Count` = sum(count))
#  Graph the new `sum_count` variable
ggplot(sum_cycle_count, aes(x = Date, y = `Total Count`)) +
  geom_line(colour = "blue") +
  theme_minimal() +
  scale_x_date(date_labels = "%b-%Y") +
  scale_y_continuous(labels = comma)
```

We can also use the latitude and longitude columns to plot the points on
a map.

```{r, fig.width=5}
library(sf)
# Convert data frame into 'sf' geospatial dataframe
cycle_counts_geo <- st_as_sf(cycle_count, coords = c("longitude", "latitude"))
# Sum counts by location
cycle_counts_geo_sum <- cycle_counts_geo %>%
  group_by(location) %>%
  summarise(`Total Cycle Count` = sum(count))
# plot
c <- ggplot(cycle_counts_geo_sum) +
  geom_sf(aes(size = `Total Cycle Count`), colour = "blue", fill = NA) +
  theme_minimal()
c
```

Can you guess which points in Glasgow are most popular for cyclists? We
can add a base layer map to reveal the streets.

```{r, message=FALSE, fig.width=7}
library(ggmap)
library(purrr)

# Create bounding box to download background street map layer
bbox_glasgow <- map_dbl(st_bbox(cycle_counts_geo_sum), 1)
names(bbox_glasgow) <- c("left", "bottom", "right", "top")

# Download background layer
glasgow <- suppressMessages(ggmap(get_stamenmap(bbox_glasgow,
  zoom = 11,
  https = TRUE,
  messaging = TRUE
)))

# Plot background with cycle counts on top
cycle_map <- glasgow + geom_point(
  data = cycle_counts_geo_sum,
  aes(
    x = unlist(map(geometry, 1)),
    y = unlist(map(geometry, 2)),
    size = `Total Cycle Count`
  ),
  col = "green",
  pch = 16,
  alpha = 0.4
) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank()
  )


cycle_map
```

```{r, echo=FALSE}
# Find top location for cycle counts
loc <- cycle_counts_geo_sum %>%
  arrange(desc(`Total Cycle Count`)) %>%
  pull(location) %>%
  head(, n = 1)
```

```{r, echo=FALSE}
# Get co-ordinates of top cycle count point
coord <- cycle_counts_geo_sum %>%
  arrange(desc(`Total Cycle Count`)) %>%
  head(, n = 1) %>%
  mutate(
    long = unlist(map(geometry, 1)),
    lat = unlist(map(geometry, 2))
  ) %>%
  select(lat, long) %>%
  sf::st_drop_geometry() %>%
  tidyr::unite(col = "coord", sep = "/") %>%
  pull(coord)
```

Looks like [`r loc`](https://www.openstreetmap.org/#map=17/`r coord`) area
is a hot spot for cyclists.
