---
title: "Riding a Bike"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Riding a Bike}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 5,
  echo = FALSE
)
```

Let's take a look at the Glasgow bike hire trip data. Firstly, load all the libraries we'll need for the analysis. Note, you may need to install these first.

```{r setup, echo=TRUE, message=FALSE}
# install.packages("devtools")
# devtools::install_github("fozy81/opendatascot")
library(opendatascotland)
library(sf)
library(tidyverse)
library(ggmap)
library(hms)
library(lubridate)
library(knitr)
```

### Download Meta-data

Search [opendata.scot](opendata.scot) for meta-data about Glasgow Next Bike Cycle Hire trips, this describes the data and who provides it. Also we can confirm which open data licence is being used.

```{r, echo = TRUE, comment=">>"}
# Search and download
metadata <- search_ods(search = "Next Bike Cycle Hire - Glasgow")
metadata$licence
```

### Download

Using the meta-data, we can download the raw cycle trips data directly from the provider and save it to our computer for future (and quick) reference.

```{r, echo=FALSE, message=FALSE}
trips <- get_ods(metadata, ask = FALSE)
# Data is returned as a list of data frames, extract first data frame in the
# list.
trips <- trips[[1]]
```

``` {r
trips <- get_ods(metadata)
# Data is returned as a list of data frames, extract first data frame in the
# list.
trips <- trips[[1]]
```

### Summary

Here we'll get some summary stats to help us understand the data.

```{r, echo = TRUE}
median_duration <- paste(round(median(trips$Duration) / 60), "mins")

```

Median trip duration is around: `` `r median_duration` ``

Top 3 start locations:

```{r, echo = TRUE}
 trips %>% 
  group_by(`Rental place`) %>% 
  summarise(count = n()) %>% 
   arrange(-count) %>% 
   head(3) %>% 
   kable()
```

Top 3 return locations:

```{r, echo = TRUE}
 trips %>% 
  group_by(`Return place`) %>% 
  summarise(count = n()) %>% 
   arrange(-count) %>% 
   head(3) %>% 
   kable()
 
```

### Average start time over 24hrs

Below is a plot of trip start times over 24hrs. Looks as expected: Peaks in morning rush hours 7.30-9.30 and evening around 6pm. Quiet at 4am.

```{r, echo = TRUE, warning=FALSE, message=FALSE}
# Start time of day (hour)
trips$hour <- hour(trips$`Start time`)

# ggplot bar chart
hour_plot <- trips %>% 
  ggplot(aes(x = hour)) +
  geom_bar(aes(fill = ..count..)) +
  scale_fill_gradient(low = "pink", high = "blue") +
  theme_minimal()

hour_plot
```

### Annual monthly usage

We can see significant usage in winter but almost a doubling of usage in summer months. Although, usage in 2022 looks a little unusual.

```{r,  echo = TRUE, warning=FALSE, message=FALSE}
# Extract year (as character to help plot as discrete categories in chart)
trips$year <- as.character(year(trips$`Start time`))
# Extract month as abbreviated 'Aug'...
trips$month <- month(trips$`Start time`, abbr = TRUE, label = TRUE)

# Plot bar chart
month_plot <- trips %>% 
  ggplot(aes(x = month)) +
  geom_bar(aes(fill = year)) +
  theme_minimal()

month_plot
```

### Map trip start points

Here we plot the start location using background map data from [OpenStreetMap](https://www.openstreetmap.org/copyright). We can see bike hire trips being used widely across the city. There's some strange lat/lon start points which need to be filtered out. Not sure what explains these but filter to a bounding box around Glasgow helps.

```{r, echo = TRUE, fig.height = 3, warning=FALSE, message=FALSE}
# Clean out weird points:
# hist(trips$`START LAT`)
trips <- trips %>%
  filter(!is.na(`START LAT`) &
    !is.na(`START LONG`) &
  `START LAT` < 55.9 &
  `START LAT` > 55.8 &
  `START LONG` < -4.0 &
  `START LONG` > -4.4)

# Convert data frame into 'sf' geospatial dataframe
trips <- st_as_sf(trips, coords = c("START LONG", "START LAT"))

# Bounding box for background map
bbox <- map_dbl(st_bbox(trips), 1)
names(bbox) <- c("left", "bottom", "right", "top")

# Download background map using bounding box
map <- suppressMessages(ggmap(get_stamenmap(bbox, zoom = 12)))

# Plot background with trips start points on top
cycle_map <- map + geom_point(
  data = trips,
  aes(
    x = unlist(map(geometry, 1)),
    y = unlist(map(geometry, 2))
  ),
  colour = "green",
  pch = 16,
  alpha = 0.6
) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank()
  )

cycle_map
```
