---
title: "Riding a Bike"
editor_options: 
  chunk_output_type: inline
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 5,
  echo = FALSE
)
```

Let's take a look at the Glasgow bike hire trip data. Firstly, load all the libraries we'll need for the analysis. Note, you may need to install these first.

```{r setup, echo=TRUE, message=FALSE}
# install.packages("devtools")
# devtools::install_github("fozy81/opendatascot")
library(opendatascotland)
library(sf)
library(tidyverse)
library(ggmap)
library(hms)
library(data.table)
library(lubridate)
library(knitr)
library(sfheaders)
library(gganimate)
library(transformr)
```

### Download Meta-data

Search [opendata.scot](opendata.scot) for meta-data about Glasgow Next Bike Cycle Hire trips, this describes the data and who provides it. Also we can confirm which open data licence is being used.

```{r, echo = TRUE, comment=">>"}
# Search and download
metadata <- search_ods(search = "Next Bike Cycle Hire - Glasgow")
metadata$licence
```

### Download

Using the meta-data, we can download the raw cycle trips data directly from the provider and save it to our computer for future (and quick) reference.

```{r, echo=FALSE, message=FALSE}
trips <- get_ods(metadata, ask = FALSE)
# Data is returned as a list of data frames, extract first data frame in the
# list.
trips <- trips[[1]]
```

``` {r
trips <- get_ods(metadata)
# Data is returned as a list of data frames, extract first data frame in the
# list.
trips <- trips[[1]]
```

### Summary

Here we'll get some summary stats to help us understand the data.

```{r, echo = TRUE}
median_duration <- paste(round(median(trips$Duration) / 60), "mins")
```

Median trip duration is around: `` `r median_duration` ``

Top 3 start locations:

```{r, echo = TRUE}
trips %>%
  group_by(`Rental place`) %>%
  summarise(count = n()) %>%
  arrange(-count) %>%
  head(3) %>%
  kable()
```

Top 3 return locations:

```{r, echo = TRUE}
trips %>%
  group_by(`Return place`) %>%
  summarise(count = n()) %>%
  arrange(-count) %>%
  head(3) %>%
  kable()
```

### Average start time over 24hrs

Below is a plot of trip start times over 24hrs. Looks as expected: Peaks in morning rush hours 7.30-9.30 and evening around 6pm. Quiet at 4am.

```{r, echo = TRUE, warning=FALSE, message=FALSE}
# Start time of day (hour)
trips$hour <- hour(trips$`Start time`)

# ggplot bar chart
hour_plot <- trips %>%
  ggplot(aes(x = hour)) +
  geom_bar(aes(fill = ..count..)) +
  scale_fill_gradient(low = "pink", high = "blue") +
  theme_minimal()

hour_plot
```

### Annual monthly usage

We can see significant usage in winter but almost a doubling of usage in summer months. Although, usage in 2022 looks a little unusual.

```{r,  echo = TRUE, warning=FALSE, message=FALSE}
# Extract year (as character to help plot as discrete categories in chart)
trips$year <- as.character(year(trips$`Start time`))
# Extract month as abbreviated 'Aug'...
trips$month <- lubridate::month(trips$`Start time`, abbr = TRUE, label = TRUE)

# Plot bar chart
month_plot <- trips %>%
  ggplot(aes(x = month)) +
  geom_bar(aes(fill = year)) +
  theme_minimal()

month_plot
```

### Map trip start points

Here we plot the start location using background map data from [OpenStreetMap](https://www.openstreetmap.org/copyright). We can see bike hire trips being used widely across the city. There's some strange lat/lon start points which need to be filtered out. Not sure what explains these but filter to a bounding box around Glasgow helps.

```{r, echo = TRUE, fig.height = 3, warning=FALSE, message=FALSE}
# Clean out weird points:
# hist(trips$`START LAT`)
trips <- trips %>%
  filter(!is.na(`START LAT`) &
    !is.na(`START LONG`) &
    `START LAT` < 55.9 &
    `START LAT` > 55.8 &
    `START LONG` < -4.0 &
    `START LONG` > -4.4)
```

```{r}
# Convert data frame into 'sf' geospatial dataframe
trips <- st_as_sf(trips, coords = c("START LONG", "START LAT"))
# trips <- select(trips, year)
# trips <- as.numeric(trips$year)

# Bounding box for background map
bbox <- map_dbl(st_bbox(trips), 1)
names(bbox) <- c("left", "bottom", "right", "top")

# Download background map using bounding box
map <- suppressMessages(ggmap(get_stamenmap(bbox, zoom = 12)))

# Plot background with trips start points on top
cycle_map <- map + geom_point(
  data = trips,
  aes(
    x = unlist(map(geometry, 1)),
    y = unlist(map(geometry, 2))
  ),
  colour = "hotpink1",
  pch = 16,
  size = 0.5,
  alpha = 0.3
) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank()
  ) + facet_wrap(vars(year))


cycle_map
```

## Start and End trip line
```{r, lines}
trips <- get_ods(metadata)
trips <- trips[[1]]
trips <- trips %>%
  filter(!is.na(`START LAT`) &
    !is.na(`START LONG`) &
    !is.na(`END LAT`) &
    !is.na(`END LONG`) &
    `START LAT` < 55.9 &
    `START LAT` > 55.8 &
    `START LONG` < -4.1 &
    `START LONG` > -4.4 &
    `END LAT` < 55.9 &
    `END LAT` > 55.8 &
    `END LONG` < -4.1 &
    `END LONG` > -4.4 &
    `END LONG` != `START LONG` &
    `END LAT` != `START LAT`)


trips <- arrange(trips, `Start time`)
trips$week <- week(trips$`Start time`)
trips$month <-  month(trips$`Start time`)
trips$year <-  year(trips$`Start time`)
trips$month  <- str_pad(trips$month, pad = 0, width = 2)
trips$week  <- str_pad(trips$week, pad = 0, width = 2)
trips$year_month <- as.integer(paste0(trips$year, trips$month))
trips$year_week <- as.integer(paste0(trips$year, trips$week ))



trips <- mutate(trips, arrival_month = 1:nrow(trips))
# trips$arrival_month <- trips$time
trips <- rename(trips,
  start_long = `START LONG`,
  start_lat = `START LAT`,
  finish_long = `END LONG`,
  finish_lat = `END LAT`
)

data <- trips

setDT(data)
data[, line_id := .I] ## Each row is a line

## create a long-form of the data
dt_line <- rbindlist(
  list(
    data[, .(arrival_month, line_id, lon = start_long, lat = start_lat, sequence = 1)],
    data[, .(arrival_month, line_id, lon = finish_long, lat = finish_lat, sequence = 2)] 
  )
)

setorder(dt_line, line_id, sequence)

sf <- sfheaders::sf_multilinestring(
  obj = dt_line,
  x = "lon",
  y = "lat",
  multilinestring_id = "arrival_month",
  linestring_id = "line_id",
  keep = T
)

sf::st_crs(sf) <- 4326 ## Assume it's in Web Mercator

print(paste0("Max distance: ",
             round(max(st_length(sf)))))
print(paste0("Min distance: ",
             round(min(st_length(sf)), 2)))
print(paste0("Average distance: ", 
             round(mean(st_length(sf)))))



# 
# g <- ggplot() +
#   geom_sf(data = sf, mapping = aes(colour = arrival_month), show.legend = FALSE) +
#   lims(x = c(4.4, 4.1), y = c(55.8, 55.9)) +
#   labs(title = 'Year-Month: {frame_time}')
# 
# g + transition_time(arrival_month) +
#     shadow_mark(alpha = 0.3, size = 0.5) 

# anim_save()

steps <- seq(from = 1000, to = 664586, by = floor(nrow(sf) / 100))
sf$year <- year(trips$`Start time`)
names(steps) <- 1:length(steps)
save_images <- map(steps, function(step) {
  n <- steps[steps == step]
  data <- sf[1:step, ]
  g <- ggplot() +
  geom_sf(data = data, mapping = aes(colour = arrival_month, 
                                     alpha = 0.9), 
          show.legend = FALSE) +
  lims(x = c(4.4, 4.1), y = c(55.8, 55.9)) +
    ggtitle(max(data$year)) +
    theme_void() +
      theme(plot.background = element_rect(fill = "black", color = "black"), 
            plot.title = element_text(color="purple",
                                      size=14, 
                                      face="bold",
                                      hjust = 0.5,
                                      vjust = 0
                                      )) +
      scale_colour_gradient(low = "blue", high = "violet",trans = "reverse") 
    
  ggsave(paste0("images/", names(n), ".png"))
})


# ffmpeg -i video.mp4 -r 12 -s 320x180 output.gif
# ffmpeg -f image2 -framerate 15 -i 'images/%d.png' -vcodec libx264 -crf 22 video.mp4
# library(animation)
# imgs <- list.files(path = "/home/pop/Documents/projects/opendatascot/vignettes/images" ,
#                    pattern="*.png")
# saveVideo({
#   for(img in imgs){
#     im <- magick::image_read( path = "/home/pop/Documents/projects/opendatascot/vignettes/images" ,
#       img)
#     plot(as.raster(im))
#   }  
# })


```


## Routes

```{r}
# 
# install.packages("geojsonsf")
# library(geojsonsf)
# loop through sf and find a route from osrm...
trips$id <- 1:nrow(trips)
steps <- seq(from = 1000, to = 664586, by = floor(nrow(trips) / 500))

purrr::map_df(split(trips, trips$id), function(trip) {
  message(trip$id)
  url <- paste0("http://127.0.0.1:5000/route/v1/bicycle/",
 trip$start_long, ",", trip$start_lat, ";", trip$finish_long, ",", trip$finish_lat, "?steps=false&geometries=geojson&exclude=motorway")
  route <- readLines((curl::curl(url)))
 # route <-  readLines(curl::curl("http://router.project-osrm.org/route/v1/driving/13.388860,52.517037;13.397634,52.529407?exclude=[motorway]"))
 
route <-  jsonlite::fromJSON(route)
json_route <- jsonlite::toJSON(route$routes$geometry)
# Need to remove '[' bracket at start and end of json. 
json_route <- substr(json_route, start=2, stop = nchar(json_route)-1)
geo <- st_read(json_route, quiet = TRUE)
st_write(geo, "trips.geojson", append = TRUE, quiet = TRUE)
  
})



# steps <- seq(from = 1, to = 1000,
#              by = floor(nrow(map_routes) / 20))
# names(steps) <- 1:length(steps)
# n <- 1
trips$day <- day(trips$`Start time`)
trips$day  <- str_pad(trips$day, pad = 0, width = 2)
trips$year_day <- as.integer(paste0(trips$year, trips$week , trips$day))

map_routes3 <- bind_cols(routes, trips[1:10000,])
map_routes3$lat_1 <- seq(from = -4.29202, to = -4.4, by = -((-4.26202 - -4.4) / 4000))[1:4000]

map_routes3$lat_2 <- seq(from = -4.21866, to = -4.1, by = -((-4.24866 - -4.1) / 4000))[1:4000]
map_routes3$lon_3 <- seq(from = 55.8347, to = 55.8, by = -((55.8647 - 55.7) / 4000))[1:4000]
map_routes3$lon_4 <- seq(from = 55.8966, to = 55.9, by = -((55.8666 - 55.9) / 4000))[1:4000]
# map_routes <- bind_rows(map_routes, map_routes)
steps <- trips[1:4000,] %>% group_by(year_day) %>%
  summarise(steps = max(id))
   steps$id <- 1:nrow(steps)       
plots <- map(split(steps, steps$id), function(step) {
  # browser()
   n <- step$id
   data <- map_routes3[1:step$steps, ]
   lat_1 <- min(data$lat_1) 
   lat_2 <- max(data$lat_2) 
   lon_1 <- min(data$lon_3) 
   lon_2 <- max(data$lon_4) 
     # data <- map_routes3
 
  ggsave(paste0("routes/", n, ".png"), width = 5, height = 4) 
  # return(g)
  
})


n <- 1
aligned_plots <- align_plots(plots[[1]],plots[[23]] ,align="hv")

map(aligned_plots[[2]], function(map) {
  

  ggsave(paste0("routes/", n, ".png"), width = 5, height = 4)
  n <<- n +1
})


# curl "https://router.project-osrm.org/route/v1/bike/13.388860,52.517037;13.385983,52.496891?steps=false&geometries=geojson"
#  curl "http://127.0.0.1:5000/route/v1/bike/13.388860,52.517037;13.385983,52.496891?steps=false&geometries=geojson"
```

